The LBRY Trending Algorithm

### 2021-11-18 tex/ms.tex
More sections written

\subsection{Interpolation}
The two ideas $M_1()$ and $M_2()$ can be merged into a compromise third option,
as follows:
\begin{align}
M_3(\ell_t, \ell_{t+1}) &= M_1(\ell_t, \ell_{t+1})^\alpha
                           M_2(\ell_t, \ell_{t+1})^{1 - \alpha}.
\end{align}
This is linear interpolation of $\log M$ values, or alternatively, it is
`geometric interpolation'. If $\alpha$ is close to 1 then the rule is more
like $M_1$, if it is close to zero it is more like $M_2$.

@@

@@

The formula for updating a trending score is Equation~\ref{eqn:decay_with_spike}:
\begin{align}
Y_{t+1} &= kY_t + M(\ell_t, \ell_{t+1}).
\end{align}
However, instead of storing $Y$ for each claim, we could instead store
a transformed version of it:
\begin{align}
Z_t &= Y_t k^{-t}.
\end{align}
The updating formula for $Y$ can then be rewritten as
\begin{align}
Z_{t+1} k^{t+1} &= kZ_t k^t + M(\ell_t, \ell_{t+1}).
\end{align}
Solving for $Z_{t+1}$ gives
\begin{align}
Z_{t+1} &= kZ_t k^t k^{-(t+1)} + M(\ell_t, \ell_{t+1})k^{-(t+1)} \\
        &= Z_t + M(\ell_t, \ell_{t+1})k^{-(t+1)}.
\end{align}
Thus, if we maintain $Z$ instead of $Y$, 
we replace `exponential decay' with maintenance of a constant trending
score for all untouched claims --- the formula reduces to $Z_{t+1}=Z_t$.
As time goes on, the mass of applied spikes
merely grows exponentially (recall that $k$ is just below 1, so $k$
raised to a negative power becomes large and positive),
and that is how recent changes become more significant than old ones.

@@

\section{The `Squashed' Grade}
Suppose $x$ is a huge number, where double precision overflow might occur if
we were to use it in a calculation (this happens at around $10^{308}$).
If $x$ is positive, we could instead deal with logarithms. Then, to add
two huge numbers without overflow, we could use
\begin{align}
\log(x + y)  &= \log(\exp(\log x) + \exp(\log y)) \\
             &= \logsumexp(\log x, \log y),
\end{align}
where $\logsumexp$ is given by
\begin{align}
\logsumexp(a, b) &= \max(a, b)
        + \log\left[\exp(a - \max(a, b)) + \exp(b - \max(a, b))\right].
\end{align}.
Subtracting the maximum before doing the exponentiation prevents overflow.
Similarly, differences could be computed on the log scale. Assuming $x>y$,
we can use the following method:
\begin{align}
\log(x - y)  &= \log(\exp(\log x) - \exp(\log y)) \\
             &= \logdiffexp(\log x, \log y),
\end{align}
where $\logdiffexp$ is given by
\begin{align}
\logdiffexp(a, b) &= x
        + \log\left[1.0 - \exp(y - x)\right].
\end{align}.

However,
logarithms only work for positive $x$, and our trending scores can be
negative (and zero) as well. In principle, any monotonic function will work
(and will allow the same total ordering of claims),
but we need one where arithmetic is possible without overflow.

Instead of $\log$, define a squashing function $S$ as:
\begin{align}
S(x) &=
    \left\{
        \begin{array}{lr}
            \log (1 + x), & \textnormal{if } x \geq 0, \\
            -\log(1 - x), & \textnormal{otherwise}.
        \end{array}
    \right.
\end{align}
The inverse, unsquashing function, is
\begin{align}
U(x) = S^{-1}(x) &=
    \left\{
        \begin{array}{lr}
            \exp(x) - 1,  & \textnormal{if } x \geq 0, \\
            1 - \exp(-x), & \textnormal{otherwise}.
        \end{array}
    \right.
\end{align}

To compute updated trending scores, we need to add and multiply
values, but on the
squashed grade. In other words, if we have squashed values $x$ and
$y$, we will need the squashed sum
\begin{align}
x \oplus y &= S(U(x) + U(y))
\end{align}
and the squashed product
\begin{align}
x \otimes y &= S(U(x)U(y)).
\end{align}
These can be worked out in terms of logsumexp and logdiffexp.
The results are fairly complicated and are given in the Appendix.

@@

>>>>

\subsection{Interpolation}
The two ideas $M_1()$ and $M_2()$ can be merged into a compromise third option,
as follows:
\begin{align}
M_3(\ell_t, \ell_{t+1}) &= M_1(\ell_t, \ell_{t+1})^\alpha
                           M_2(\ell_t, \ell_{t+1})^{1 - \alpha}. \label{eqn:spike_mass}
\end{align}
This is linear interpolation of $\log M$ values, or alternatively, it is
`geometric interpolation'. If $\alpha$ is close to 1 then the rule is more
like $M_1$, if it is close to zero it is more like $M_2$.

@@

% TODO: Consider idea of symmetrising by using both old and new LBC values,
% as this may make reorg safety occur naturally without any extra effort

@@

The formula for updating a trending score is Equation~\ref{eqn:decay_with_spike}:
\begin{align}
Y_{t+1} &= kY_t + M(\ell_t, \ell_{t+1}).
\end{align}
However, instead of storing $Y$ for each claim, we could instead store
a transformed version of it:
\begin{align}
Z_t &= Y_t k^{-t}.
\end{align}
The updating formula for $Y$ can then be rewritten as
\begin{align}
Z_{t+1} k^{t+1} &= kZ_t k^t + M(\ell_t, \ell_{t+1}).
\end{align}
Solving for $Z_{t+1}$ gives
\begin{align}
Z_{t+1} &= kZ_t k^t k^{-(t+1)} + M(\ell_t, \ell_{t+1})k^{-(t+1)} \\
        &= Z_t + M(\ell_t, \ell_{t+1})k^{-(t+1)}. \label{eqn:sparse_update}
\end{align}
Thus, if we maintain $Z$ instead of $Y$, 
we replace `exponential decay' with maintenance of a constant trending
score for all untouched claims --- the formula reduces to $Z_{t+1}=Z_t$.
As time goes on, the mass of applied spikes
merely grows exponentially (recall that $k$ is just below 1, so $k$
raised to a negative power becomes large and positive),
and that is how recent changes become more significant than old ones.

@@

\section{The `Squashed' Grade}
Suppose $x$ is a huge number, where double precision overflow might occur if
we were to use it in a calculation (this happens at around $10^{308}$).
If $x$ is positive, we could instead deal with logarithms. Then, to add
two huge numbers without overflow, we could use
\begin{align}
\log(x + y)  &= \log(\exp(\log x) + \exp(\log y)) \\
             &= \logsumexp(\log x, \log y),
\end{align}
where $\logsumexp()$ is given by
\begin{align}
\logsumexp(a, b) &= \max(a, b)
        + \log\left[\exp(a - \max(a, b)) + \exp(b - \max(a, b))\right].
\end{align}
Subtracting the maximum before doing the exponentiation prevents overflow.
Similarly, differences could be computed on the log scale. Assuming $x>y$,
we can use the following method:
\begin{align}
\log(x - y)  &= \log(\exp(\log x) - \exp(\log y)) \\
             &= \logdiffexp(\log x, \log y),
\end{align}
where $\logdiffexp$ is given by
\begin{align}
\logdiffexp(a, b) &= a
        + \log\left[a - \exp(b-a)\right].
\end{align}
However,
logarithms only work for positive $x$, and our trending scores can be
negative (and zero) as well. In principle, any monotonic function will work
(and will allow the same total ordering of claims),
but we need one where arithmetic is possible without overflow.

Instead of $\log$, define a squashing function $S$ as:
\begin{align}
S(x) &=
    \left\{
        \begin{array}{lr}
            \log (1 + x), & \textnormal{if } x \geq 0, \\
            -\log(1 - x), & \textnormal{otherwise}.
        \end{array}
    \right.
\end{align}
The inverse, unsquashing function, is
\begin{align}
U(x) = S^{-1}(x) &=
    \left\{
        \begin{array}{lr}
            \exp(x) - 1,  & \textnormal{if } x \geq 0, \\
            1 - \exp(-x), & \textnormal{otherwise}.
        \end{array}
    \right.
\end{align}
For large positive values, these are approximately equal to log and exp
respectively, but for smaller values (and negative values), it remains
continuous.

To compute updated trending scores, we need to add and multiply
values using Equations~\ref{eqn:sparse_update}~and~\ref{eqn:spike_mass}, but on the
squashed grade. In other words, if we have squashed values $x$ and
$y$, we will need the squashed sum
\begin{align}
x \oplus y &= S(U(x) + U(y))
\end{align}
and the squashed product
\begin{align}
x \otimes y &= S(U(x)U(y)).
\end{align}
These can be worked out in terms of logsumexp and logdiffexp.
The results are fairly complicated and are given in the Appendix.

@@

Squashed multiplication is simpler:
\begin{align}
x \otimes y &= \sign(xy)
                \logsumexp\left(
                    \logdiffexp(|x|, 0) + \logdiffexp(|y|, 0),
                0\right).
\end{align}
